---
title: Climate-resilient Reclamation (CRR) Shiny App
author: Peter Solymos / Analythium Solutions
date: 2025-03-31
output: pdf_document
fontfamily: mathpazo
fontsize: 11pt
urlcolor: blue
---

```{r include=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
library(terra)
```

## Introduction

The Climate-resilient Reclamation (CRR) Shiny App was originally developed by
Ivan Bjelanovic to help visualize baseline and future ecosystem states
in Alberta developed by the Canadian Forest Service (Natural Resources Canada).

This original version of the app was deployed to the Shinyapps.io platform and
referenced raster layers hosted on a 3rd party geospatial server.
This server became defunct and as a result the Shiny app stopped working.

The current project's goal is to revitalize the app and find sustainable and
easy-to-manage hosting option for the app and raster layers:

1. Update existing app built by Ivan Bjelanovic to current standards and packages 
   (including re-establishing broken links);
2. Add results from new Alberta simulations to be provided by ApexRMS 
   (up to 64 scenarios);
3. Provide commented R code and instructions for updating the app with new layers.

This report outlines how the 3 delivery points were accomplished.

## Project organization

The CRR Shiny App project is hosted on GitHb as a public repository:
<https://github.com/psolymos/crr-shiny>. 
To get the source code, visit the link and click green "Code" button to download
a zip file with the code in it. ALternatively, you can use `git` to clone
the repository with `git clone https://github.com/psolymos/crr-shiny.git`.

The repository has the following folder:

- `app-v1`: Original (v1) Shiny app with updated references to static files.
- `app-v2`: V2 of the Shiny app with flexibility for adding new layers.
- `data`: Example raster files and R scripts for data processing.
- `docs`: Folder with static files to be served through GitHub Pages.

## App deployment

The app needs 2 different deployments:

1. The Shiny app that is responsible for the user interface and the
   application logic;
2. and the file server that hosts the spatial files that are too big
   and numerous for being deployed alongside the Shiny app.

Shiny apps can be hosted in many ways. The easiest and free option is to
register for a Shinyapps.io account at <https://www.shinyapps.io/auth/oauth/signup>.
The free account's credentials can be added to RStudio Desktop.
Then, the app can be deployed from RStudio within a user account.

### Legacy app (v1)

The original version of the Shiny app was updated the following ways:

- R package dependencies revised and updated to reflect changes around the
  geospatial package ecosystem (retiring `sp`, `geos`, etc. in favor of `sf`,
  `terra` and others);
- Updated links to external file server (see below the _File server_ section).

The source code of the updated version of the original app is in the `app-v1`
folder of the GitHub repository.
The current live deployment of this v1 of the app can be found at:
<https://analythium.shinyapps.io/crr-shiny-v1/>.

Install dependencies (from inside the `app-v1` folder) using the 
`dependencies.json` file:

```R
if (!requireNamespace("deps")) install.packages("deps")
deps::install()
```

To edit the v1 of the app, double click the Rproj file in the `app-v1`
folder -- this will open the project in RStudio Desktop.

To run the app, click on the "play" button in RStudio,
call `shiny::runApp("app-v1")` from R or
`R -q -e 'shiny::runApp("app-v1")'` from shell.

### Updated Shiny app (v2)

The v1 of the Shiny app has the following updates compared to the v1:

- It is a stripped down version focusing on scenario comparisons;
- It is set up for easier maintenance explained later in this document;
- It has fewer dependencies.

The source code of the updated version of the original app is in the `app-v2`
folder of the GitHub repository.
The current live deployment of this v2 of the app can be found at:
<https://analythium.shinyapps.io/crr-shiny-v2/>.

Install dependencies (from inside the `app-v2` folder) with:

```R
if (!requireNamespace("deps")) install.packages("deps")
deps::install()
```

This command will use the `dependencies.json` file to install missing R packages.

To edit the v1 of the app, double click the Rproj file in the `app-v2`
folder -- this will open the project in RStudio Desktop.

To run the app, click on the "play" button in RStudio,
call `shiny::runApp("app-v2")` from R or
`R -q -e 'shiny::runApp("app-v2")'` from shell.

### File server

The links in the Shiny apps point to the GitHub Pages based file server that is 
part of the `docs` folder. GitHub pages are available for free of charge for
public GitHub repositories. It needs to be enabled in the Settings/Pages.
The repository serves static files from the `docs` folder of the main Git branch.
When new files are added to the repository, the files will be published
on the `github.io` website.

E.g. for this setup, the file `<github-repository-url>/docs/index.html` will 
be published at `https://<github-pages-url/index.html`.
The file `https://github.com/psolymos/crr-shiny/docs/v1/scenarios/...tif`
will be available at `https://psolymos.github.io/crr-shiny/docs/v1/scenarios/...tif`.

GitHub pages can be accessed from other websites because it allows
cross origin resource sharing (via CORS headers). This makes this service
ideal for hosting static assets. In our case, GeoTIFF files.

### Collaborative code management

The repository is hosted under the `psolymos` personal GitHub account.
The repository is public, which means that others can fork the project.
Edits to forked projects can be merged to the upstream repository
after submitting a pull request. But the changes do not have to be merged
for others to develop the project and create independent Shinyapps.io
deployments. However, submitting pull requests is good practice.

The owner of the repository (`psolymos`) can grant write access to other
GitHub users. If that is desired, please let me know.

The ownership of the repository can also be transferred to another user or
GitHub organization.

## Adding new layers

The v2 of the Shiny app is developed with ease of maintenance in mind.
Here we outline how to process new raster layers and add them to the app
and the file server.

### Raster layer processing

A new layer is expected to be a GeoTIFF file with a single band, that band
should have positive integers as values, or missing. The cartographic 
reference system needs to be know, otherwise assumed to be 
Alberta 10-TM (Forest) (EPSG:3400). The raster cell resolution is
500 m by 500 m.

The cell values denote ecotype classes, such as:

```{r echo=FALSE}
es <- read.csv("ecosite_classes.csv")
knitr::kable(es)
```

An example input file is `Stralberg_baseline_vegeco4top_mod_int.tif`:

```{r}
library(terra)
r1 <- rast("Stralberg/Stralberg_baseline_vegeco4top_mod_int.tif")
r1
```

We can set the color table for the ecosite classes as:

```{r}
es <- read.csv("ecosite_classes.csv")
ct <- data.frame(value=es$Code, col=es$Color)
str(ct)

coltab(r1) <- ct
```

We can now plot the raster and get the assigned color mapping:

```{r}
plot(r1)
```

#### Saving integer raster

To save the integer version in a compressed format and without loss of information,
use the following command:

```R
writeRaster(r1, "file_int.tif", overwrite=TRUE, datatype="INT1U")
```

This will use the default LZW compression (lossless) and unsigned 8-bit integer.
This can store values between 0 and 254 (2^8-1, minus one for `NA` storage), 
raster values are rounded if not already integer.
This will result in the smallest file size (<1Mb for a 500 m resolution raster
for Alberta).

The integer version can be further processed and and summarized.

#### Save RGB

The colorized (RGB: red-green-blue) version is useful to show the cells
as the desired color without having to use a color table. This is important
because Leaflet maps can't easily display categorical raster layers
without extensive JavaScript coding.

The colorized version is also reprojected to web mercator that is the 
standard for web mapping:

- EPSG:4326 is used for feature geometries (i.e. roads, etc.),
- EPSG:3857 is used for base layers and similar (raster, image) objects.

We will use an existing TIF file as the template that has the
desired EPSG3857 projection and matching resolution to our integer map:

```{r}
rt <- rast("Stralberg/Stralberg_baseline_vegeco4top_mod_prj_on.tif")
rt
```

We need to use the nearest neighbor method:

```{r}
r1w <- project(r1, rt, method = "near")
```

Now we can colorize the layer as:

```{r}
r1c <- colorize(r1w, "rgb", alpha=FALSE)
```

```{r}
plot(r1c)
```

Use unsigned 16-bit integer raster format for the RGB image to avoid
potential issues with the missing data mask (alpha channel).
Use the `DEFLATE` compression method for best results:

```R
writeRaster(r1c, "file_rgb.tif", overwrite=TRUE, datatype="INT2U", gdal="COMPRESS=DEFLATE")
```

The output file is usually 1.5-2 times larger than integer version.

#### ApexRMS provided rasters

The raster values provided by ApexRMS (`EcositeStateID` column) for the new 
scenarios need to be mapped to the ecosite classes using a lookup table:

```{r echo=FALSE}
lt <- read.csv("ecosite_classes_lookup.csv")
knitr::kable(lt)
```

Here is an example raster file:

```{r}
r2 <- rast("STSim_Outputs/sc.it1.ts100.tif")
r2
summary(r2)
```

We can apply the ecosite classification as follows:

```{r}
vo <- values(r2) # old values
summary(vo)

vn <- vo         # new values
lt <- read.csv("ecosite_classes_lookup.csv")
vn[,1] <- ifelse(is.na(vo[,1]), NA, lt$Code[match(vo[,1], lt$EcositeStateID)])
summary(vn)

r3 <- r2         # create a copy
values(r3) <- vn # set values
```

We can now apply the previous color table to this new raster:

```{r}
coltab(r3) <- ct
plot(r3)
```

Saving the integer version, projecting to web mercator and colorization,
finally saving the RGB version is the same as shown before.

### Adding the layer to the file server

The layer can now be added to the `docs` folder so it is served via GitHub
pages. The `docs` folder has the following high level structure:

- `testing`: folder for testing related files,
- `v1`: files served for the v1 of the Shiny app,
- `v2`: this will be used for organizing future layers for the v2 of the Shiny app.

The `v1` folder currently contains the following files:

```bash
docs/v1
|-- base
|   |-- BaselineUpland_prj_on.tif
|   |-- DEP_baselayer_on.tif
|   |-- PLC30_OS_100_prj_on.tif
|   +-- Stralberg_baseline_vegeco4top_mod_prj_on.tif
+-- scenarios
    |-- climate-driven
    |   |-- Stralberg_20112040_CSIRO_prj_on.tif
    |   |-- Stralberg_20112040_CanESM2_prj_on.tif
    |   |-- Stralberg_20112040_HadGEM2_prj_on.tif
    |   |-- Stralberg_20412070_CSIRO_prj_on.tif
    |   |-- Stralberg_20412070_CanESM2_prj_on.tif
    |   |-- Stralberg_20412070_HadGEM2_prj_on.tif
    |   |-- Stralberg_20712100_CSIRO_prj_on.tif
    |   |-- Stralberg_20712100_CanESM2_prj_on.tif
    |   +-- Stralberg_20712100_HadGEM2_prj_on.tif
    |-- fire-mediated-constrained
    |   |-- ab_veg_fm_c_20112040_CSIRO_prj_on.tif
    |   |-- ab_veg_fm_c_20112040_CanESM2_prj_on.tif
    |   |-- ab_veg_fm_c_20112040_HadGEM2_prj_on.tif
    |   |-- ab_veg_fm_c_20412070_CSIRO_prj_on.tif
    |   |-- ab_veg_fm_c_20412070_CanESM2_prj_on.tif
    |   |-- ab_veg_fm_c_20412070_HadGEM2_prj_on.tif
    |   |-- ab_veg_fm_c_20712100_CSIRO_prj_on.tif
    |   |-- ab_veg_fm_c_20712100_CanESM2_prj_on.tif
    |   |-- ab_veg_fm_c_20712100_HadGEM2_prj_on.tif
    +-- fire-mediated-unconstrained
        |-- ab_veg_fm_uc_20112040_CSIRO_prj_on.tif
        |-- ab_veg_fm_uc_20112040_CanESM2_prj_on.tif
        |-- ab_veg_fm_uc_20112040_HadGEM2_prj_on.tif
        |-- ab_veg_fm_uc_20412070_CSIRO_prj_on.tif
        |-- ab_veg_fm_uc_20412070_CanESM2_prj_on.tif
        |-- ab_veg_fm_uc_20412070_HadGEM2_prj_on.tif
        |-- ab_veg_fm_uc_20712100_CSIRO_prj_on.tif
        |-- ab_veg_fm_uc_20712100_CanESM2_prj_on.tif
        +-- ab_veg_fm_uc_20712100_HadGEM2_prj_on.tif
```

The `base` subfolder is for baseline rasters, while the `scenarios` subfolder
has the different runs inside.

New GeoTIFF files (EPSG:3857 and RGB format) should be organized
within the `v2` folder similarly to how it was done for `v1`.
The integer version and other supporting files can also be added as needed.

### Adding the new layer to the Shiny app

The Shiny app pulls the information about the file server from a csv file.
It is in `app-v2/data/files_list.csv`:

```{r}
files_list <- read.csv("../app-v2/data/files_list.csv")
files_list$time_period <- as.character(files_list$time_period)
head(files_list)
```

This file captures the important aspects of the baseline and scenario related
raster files. The v2 app has functions that can retrieve the following info:

```R
Files$get_pred_types()
#   Climate driven   Fire mediated, constrained Fire mediated, unconstrained 
#      "Stralberg"                "ab_veg_fm_c"               "ab_veg_fm_uc" 

Files$get_time_periods()
#  2011-2040  2041-2070  2071-2100 
# "20112040" "20412070" "20712100" 

Files$get_climate_models()
#   CanESM2     CSIRO   HadGEM2 
# "CanESM2"   "CSIRO" "HadGEM2" 
```

These are used to construct the right-hand-side menu in the app.

Inside the Shiny server function, the following expressions return the
path required for the GeoTIFF URLs:

```R
Files$get_path_for_baseline("BaselineUpland_prj_on")
# [1] "v1/base/BaselineUpland_prj_on.tif"

Files$get_path_for_scenario("Stralberg", "20112040", "CanESM2")
# [1] "v1/scenarios/climate-driven/Stralberg_20112040_CanESM2_prj_on.tif"
```

If the files list CSV file reflects the file server status, the v2 Shiny app
will be able to display any new layers.

### Note about Leaflet limitations

Leaflet is a web mapping library that is used inside the Shiny app.
It uses resampling for displaying rasters at lower (coarser) zoom levels.
This behavior cannot be turned off. The benefit of the trade-off is that
the GeoTIFF layers can be displayed entirely on the client side. I.e. the 
Shiny app does not have to send into to the browser. Also, the browser can
cache the layers, so loading the 2nd time should take less time.

An alternative option is to load the integer version of the TIF from the
file server and display the color mapped or colorized version from within
the Shiny R session. This is an extremely resource intensive step and requires
patience from the user. Alberta-sized rasters can be handled this way,
and the Shiny session has to send the whole raster image to the client-side
HTML, so on top of the server side compute requirements, it is also heavy
on the client side.
